---
title: "AE 08 - Web Scraping: Books & Quotes"
author: "Your Name"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
  pdf_document:
    toc: true
    latex_engine: xelatex
    keep_tex: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, eval = FALSE)
```

## Introduction

In this application exercise, you'll learn the basics of **web scraping** - the process of extracting data from websites. You'll scrape book data from Books to Scrape and quote data from Quotes to Scrape - two sites built specifically for practicing web scraping - to create datasets you can analyze.

**Important:** Web scraping should always be done responsibly and ethically. We're using two demo sites that were built *for the purpose of* being scraped, so there are no ethical gray areas here - but you should always check a website's Terms of Service and `robots.txt` file before scraping a real site.

**In this application exercise, you will:**

-   Read HTML pages into R
-   Extract specific elements using CSS selectors
-   Clean scraped text data with string manipulation
-   Build datasets from web-scraped data
-   Manually add data by indexing into data frames
-   Practice reproducible web scraping workflows

**Estimated time:** 60-75 minutes

**Note:** This AE uses R scripts (`.R` files) instead of R Markdown for parts of the exercise. You'll work with four scripts:

1.  `01-books-scraping.R` - Scrape books from Books to Scrape (with blanks to fill)
2.  `02-quotes-scraping.R` - Scrape quotes from Quotes to Scrape (with blanks to fill)
3.  `03-books-scraping-complete.R` - Complete solution for books
4.  `04-quotes-scraping-complete.R` - Complete solution for quotes

------------------------------------------------------------------------

## Getting Started

### Fork and Clone Workflow

**Quick reminder:** Work in YOUR forked `application-exercises` repository.

1.  Navigate to the `ae-08-webscraping` folder in YOUR fork
2.  You'll see this `.Rmd` file and four `.R` script files
3.  Open `ae-08-webscraping.Rmd` (this file)
4.  Update the YAML with your name
5.  Click **Knit** to verify it works

**Working with R scripts:**

R scripts (`.R` files) are different from R Markdown (`.Rmd` files):

-   No narrative text, just code and comments
-   No knitting - you run code line by line or in chunks
-   Great for data cleaning, scraping, and automation tasks
-   Comments start with `#`

------------------------------------------------------------------------

## Web Scraping Overview

### What is Web Scraping?

**Web scraping** is the process of extracting data from websites. Instead of manually copying data, we write code to:

1.  Read the HTML of a webpage
2.  Find the specific elements we want (titles, prices, ratings, etc.)
3.  Extract the text or data from those elements
4.  Clean and organize the data into a usable format

### The `rvest` Package

We'll use the **rvest** package (pronounced "harvest") for web scraping. Key functions:

-   `read_html()`: Reads a webpage's HTML
-   `html_nodes()`: Selects specific HTML elements using CSS selectors
-   `html_text()`: Extracts text from HTML elements
-   `html_attr()`: Extracts attributes (like links or classes) from HTML elements

### How to Find CSS Selectors

CSS selectors identify specific elements on a webpage. To find them:

1.  Open the webpage in Chrome or Firefox
2.  Right-click on the element you want → "Inspect" or "Inspect Element"
3.  In the developer tools, right-click the highlighted HTML → Copy → Copy selector
4.  Or use a browser extension like SelectorGadget

**Common CSS selector patterns:**

-   `.className` - selects elements with a specific class
-   `#idName` - selects element with a specific ID
-   `tagName` - selects all elements of that type (e.g., `a` for links)
-   Combinations like `.product_pod h3 a` - selects `<a>` tags inside an `<h3>` inside class `product_pod`

------------------------------------------------------------------------

## Part 1: Scraping Books

### Overview

We'll scrape the homepage of Books to Scrape: <https://books.toscrape.com/>

This is a demo site built specifically for practicing web scraping - its own homepage even says so. It lists 20 books per page. For each book, we'll extract:

-   **Title** (e.g., "A Light in the Attic")
-   **Price** (e.g., £51.77)
-   **Star rating** (e.g., 3 stars, on a 1-5 scale)

**Note:** The prices on this site are in British pounds (£), and both the prices and star ratings were randomly assigned by the site - they don't reflect anything real about the books. That's fine for our purposes; we're practicing the *scraping technique*, not researching actual book prices.

### Exercise 1: Inspect the Webpage

Before we code, let's understand the webpage structure.

**Task 1.1:** Open <https://books.toscrape.com/> in your browser.

**Task 1.2:** Right-click on the first book's title and select "Inspect" or "Inspect Element."

**Task 1.3:** Look at the HTML structure. You'll see something like:

```{html}
<article class="product_pod">
  <p class="star-rating Three"></p>
  <h3><a href="catalogue/a-light-in-the-attic_1000/index.html" 
        title="A Light in the Attic">A Light in the ...</a></h3>
  <div class="product_price">
    <p class="price_color">£51.77</p>
  </div>
</article>
```

**What patterns do you notice?**

**Your answer:**

-   Book titles are in: \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

-   Prices are in: \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

-   Star ratings are in: \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

**Hint:** Look closely at the title on the `<a>` tag versus the visible text inside it. Also notice: the star rating isn't in any visible text at all - where does the number of stars actually live?

------------------------------------------------------------------------

### Exercise 2: Open and Complete the Books Script

**Task 2.1:** In RStudio's Files pane, open `01-books-scraping.R`

You'll see a script with blanks to fill in. The structure is:

1.  Load packages
2.  Read the HTML page
3.  Extract titles
4.  Extract prices (with string cleaning)
5.  Extract star ratings (with class-name decoding)
6.  Combine into a data frame
7.  Add rank numbers

**Task 2.2:** Work through the script and fill in ALL the blanks. Hints are provided in comments.

**Key blanks to fill:**

-   `read_html("___")` - the Books to Scrape URL
-   `html_nodes("___")` - CSS selectors for titles, prices, and star ratings
-   `html_attr("___")` - which attribute holds the *full* title (not the truncated visible text)
-   String manipulation to clean prices (remove `£`) and decode star ratings (remove `"star-rating "` from the class name, then convert the word to a number)
-   Tibble column assignments

**Task 2.3:** Run your code line by line (Ctrl+Enter or Cmd+Enter) or select all and run.

**Task 2.4:** Check your results. You should get a data frame with 20 books. Run:

```{r check-books, eval=FALSE}
# After running your script, these should work:
head(books)
nrow(books)  # Should be 20
```

**How many books did you successfully scrape?** \_\_\_\_\_\_

**What is the title of the first book on the page?** \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

------------------------------------------------------------------------

### Exercise 3: Understanding the Code

Let's break down what the books script does:

**Task 3.1:** Explain what this code does in plain English:

```{r explain-read}
page <- read_html("https://books.toscrape.com/")
```

**Your answer:**

------------------------------------------------------------------------

**Task 3.2:** Explain what the pipe `%>%` does in this code:

```{r explain-pipe}
titles <- page %>%
  html_nodes(".product_pod h3 a") %>%
  html_attr("title")
```

**Your answer:**

------------------------------------------------------------------------

**Task 3.3:** Why do we pull the `title` *attribute* instead of just using `html_text()` on the link?

**Hint:** Compare the visible text "A Light in the ..." to the `title="A Light in the Attic"` attribute in the HTML above.

**Your answer:**

------------------------------------------------------------------------

**Task 3.4:** The star rating extraction pulls a CSS *class name*, not visible text, and then has to translate a word into a number:

```{r explain-star-rating}
ratings <- page %>%
  html_nodes(".product_pod p.star-rating") %>%
  html_attr("class") %>%
  str_remove("star-rating ") %>%
  case_match(
    "One"   ~ 1,
    "Two"   ~ 2,
    "Three" ~ 3,
    "Four"  ~ 4,
    "Five"  ~ 5
  )
```

**Why does the site encode the rating in a class name (e.g., `"star-rating Three"`) instead of just writing "3 stars" as text somewhere on the page?**

**Your answer:**

------------------------------------------------------------------------

------------------------------------------------------------------------

### Exercise 4: Verify Your Data

**Task 4.1:** Check if your scraping worked correctly. Fill in the code:

```{r verify-books}
# Check structure
str(books)

# Look at the top 10
books %>%
  slice(1:10)

# Check for missing values
books %>%
  summarise(
    missing_titles  = sum(is.na(___)),
    missing_prices  = sum(is.na(___)),
    missing_ratings = sum(is.na(___))
  )
```

**How many missing values do you have?** \_\_\_\_\_\_\_

**If you have missing values, what might have gone wrong?**

------------------------------------------------------------------------

------------------------------------------------------------------------

------------------------------------------------------------------------

## Part 2: Scraping Quotes

### Overview

Now we'll scrape Quotes to Scrape: <https://quotes.toscrape.com/>

This is similar to books, but with some key differences:

-   Different CSS selectors, and one field (tags) can have a *variable number* of values per quote
-   Different HTML structure
-   We'll add additional information manually for the first 3 authors by visiting their individual bio pages

### Exercise 5: Inspect the Quotes Page

**Task 5.1:** Open <https://quotes.toscrape.com/> in your browser.

**Task 5.2:** Inspect the first quote. How is the HTML structure different from the books page?

**Your answer:**

------------------------------------------------------------------------

------------------------------------------------------------------------

**Task 5.3:** Note the CSS selectors you'll need for:

-   Quote text: \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

-   Author name: \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

-   Tags: \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

------------------------------------------------------------------------

### Exercise 6: Complete the Quotes Script

**Task 6.1:** Open `02-quotes-scraping.R` in RStudio.

**Task 6.2:** Fill in ALL the blanks in the script. This is more challenging than the books script!

**Key differences to note:**

-   The URL is different
-   The CSS selectors are different
-   Each quote can have a *different number of tags* (some have 1, some have 5), so you can't just pull "all the tags" and expect them to line up with "all the quotes" - you have to work **quote by quote**
-   You'll manually add birth date and birthplace for the first 3 *authors* by visiting their individual author pages

**Task 6.3:** For the first 3 quotes on the page, visit the linked author's individual "(about)" page and record:

**Author #1:**

-   Name: \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_
-   Born date: \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_
-   Born location: \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

**Author #2:**

-   Name: \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_
-   Born date: \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_
-   Born location: \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

**Author #3:**

-   Name: \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_
-   Born date: \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_
-   Born location: \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

**Task 6.4:** Add this information to your script using indexing:

```{r manual-data-entry, eval=FALSE}
# Example for author #1
quotes$author_born_date[1]     <- "March 14, 1879"   # Replace with actual date
quotes$author_born_location[1] <- "in Ulm, Germany"  # Replace with actual location
```

------------------------------------------------------------------------

### Exercise 7: Understanding Manual Data Entry

**Task 7.1:** Why do we use `quotes$author_born_date[1]` instead of creating a new data frame?

**Your answer:**

------------------------------------------------------------------------

------------------------------------------------------------------------

**Task 7.2:** What does the `[1]` mean in `quotes$author_born_date[1]`?

**Your answer:**

------------------------------------------------------------------------

**Task 7.3:** How would you add information for the 7th quote's author?

**Your answer:**

------------------------------------------------------------------------

------------------------------------------------------------------------

## Part 3: Comparing Approaches

### Exercise 8: Books vs Quotes - What's Different?

**Task 8.1:** Compare the two scripts. What was easier/harder in the quotes script?

**Easier in quotes:**

------------------------------------------------------------------------

**Harder in quotes:**

------------------------------------------------------------------------

------------------------------------------------------------------------

**Task 8.2:** Why did we need to loop over each quote individually (`map_chr()` per `.quote` node) to get the tags, instead of just scraping "all the tags" on the page in one shot the way we scraped "all the titles" on the books page?

**Your answer:**

------------------------------------------------------------------------

------------------------------------------------------------------------

**Task 8.3:** Looking at the complete scripts (`03-books-scraping-complete.R` and `04-quotes-scraping-complete.R`), check your work. Did you get the CSS selectors correct?

**What I got right:**

------------------------------------------------------------------------

**What I need to fix:**

------------------------------------------------------------------------

------------------------------------------------------------------------

------------------------------------------------------------------------

## Part 4: Exploring the Data

Now that you've scraped both datasets, let's analyze them!

### Exercise 9: Book Analysis

**Task 9.1:** What star rating is most common among these 20 books?

```{r rating-analysis}
books %>%
  count(rating, sort = TRUE)
```

**Answer:** A \_\_\_\_\_\_-star rating is the most common.

**Task 9.2:** What's the average price of the books on this page?

```{r avg-price}
books %>%
  summarise(avg_price = mean(___))
```

**Answer:** The average price is £\_\_\_\_\_\_

**Task 9.3:** Create a visualization showing price by star rating:

```{r price-by-rating}
ggplot(books, aes(x = ___, y = ___, group = ___)) +
  geom_boxplot() +
  labs(
    title = "___",
    x = "___",
    y = "___"
  )
```

**What pattern do you see (or not see)? Why might that make sense given how this site was built?**

------------------------------------------------------------------------

------------------------------------------------------------------------

------------------------------------------------------------------------

### Exercise 10: Quote Analysis

**Task 10.1:** How many quotes did you successfully scrape?

```{r count-quotes}
nrow(quotes)
```

**Answer:** \_\_\_\_\_\_ quotes

**Task 10.2:** Which author appears most often on this page?

```{r author-count}
quotes %>%
  count(author, sort = TRUE)
```

**Answer:** \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_ appears most often, with \_\_\_\_\_\_ quotes.

**Task 10.3:** Based on the 3 authors where you added birth information, what time periods or places seem represented?

**Your answer:**

------------------------------------------------------------------------

------------------------------------------------------------------------

------------------------------------------------------------------------

## Part 5: Reflection on Web Scraping

### Exercise 11: Challenges and Ethics

**Task 11.1:** What was the most challenging part of this exercise?

**Your answer:**

------------------------------------------------------------------------

------------------------------------------------------------------------

**Task 11.2:** Why is it important to check a website's `robots.txt` file before scraping?

**Hint:** Try visiting <https://books.toscrape.com/robots.txt> and <https://quotes.toscrape.com/robots.txt>. Both return a "Not Found" (404) page. What does that tell you about whether scraping is restricted here, versus on a real commercial site?

**Your answer:**

------------------------------------------------------------------------

------------------------------------------------------------------------

**Task 11.3:** What are some situations where web scraping might NOT be appropriate or legal?

**Your answer:**

------------------------------------------------------------------------

------------------------------------------------------------------------

------------------------------------------------------------------------

**Task 11.4:** If Books to Scrape changed their website design tomorrow, what would happen to your scraping code?

**Your answer:**

------------------------------------------------------------------------

------------------------------------------------------------------------

**Task 11.5:** Why might an API (Application Programming Interface) be preferable to web scraping?

**Your answer:**

------------------------------------------------------------------------

------------------------------------------------------------------------

------------------------------------------------------------------------

------------------------------------------------------------------------

## Part 6: Challenge (Optional)

### Exercise 12: Scrape a Second Page

**Challenge:** Books to Scrape has 50 pages of books (1,000 books total). Each page has a "next" link you can follow.

**Task 12.1:** Inspect the "next" link at the bottom of the books homepage. What CSS selector would you use to grab its `href`?

**Selector:** \_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

------------------------------------------------------------------------

**Task 12.2:** Write code to scrape page 2 as well, and combine it with your page 1 data using `bind_rows()`:

```{r challenge-scrape}
# Your code here
```

**Task 12.3:** How many total books do you now have across both pages?

**Your answer:**

------------------------------------------------------------------------

------------------------------------------------------------------------

## Knit and Submit

**Before you finish:**

1.  ✅ Make sure both R scripts run without errors
2.  ✅ Fill in all answer spaces in this .Rmd file
3.  ✅ Knit to HTML first to verify everything works
4.  ✅ Review the HTML output

**Git workflow:**

1.  Click the **Git** pane
2.  Check boxes next to ALL changed files (this .Rmd AND the .R scripts)
3.  Click **Commit**
4.  Write a commit message: "Completed AE 08 - Web Scraping"
5.  Click **Commit**
6.  Click **Push**
7.  Verify your work is on GitHub in **YOUR fork**

------------------------------------------------------------------------

## Generate PDF for Canvas Submission

Once everything works in HTML:

1.  Click the **Knit** dropdown arrow
2.  Select **Knit to PDF**
3.  Wait for PDF generation
4.  Find `ae-08-webscraping.pdf` in the Files pane
5.  Export and save to your computer
6.  Submit the PDF to Canvas

**Remember:** Submit to BOTH GitHub (your .Rmd AND .R files) AND Canvas (the PDF)!

------------------------------------------------------------------------

## Key Concepts Summary

**Web Scraping Workflow:**

1.  **Inspect** the webpage to understand HTML structure
2.  **Identify** CSS selectors for the elements you want
3.  **Read** the HTML with `read_html()`
4.  **Extract** elements with `html_nodes()` and CSS selectors
5.  **Get text or attributes** with `html_text()` / `html_attr()`
6.  **Clean** the text with stringr functions
7.  **Combine** into a data frame
8.  **Verify** your results

**Key Functions:**

-   `read_html(url)` - Read a webpage
-   `html_nodes(css)` - Select elements by CSS selector
-   `html_text()` - Extract visible text from elements
-   `html_attr(name)` - Extract an attribute (e.g., `"title"`, `"class"`, `"href"`) from elements
-   `str_remove()`, `str_remove_all()` - Remove patterns from strings
-   `str_squish()` - Remove extra whitespace
-   `case_match()` - Recode one set of values (e.g., words) into another (e.g., numbers)
-   `map_chr()` - Apply a function to each element of a list, one at a time (needed when a field like "tags" has a variable number of values per row)
-   `as.numeric()` - Convert text to numbers

**CSS Selectors:**

-   `.className` - Elements with a class
-   `#idName` - Element with an ID
-   `tag.className` - Specific tag with a class
-   `.class1 .class2` - Nested classes

------------------------------------------------------------------------

## Resources

**Learn more about web scraping:**

-   [rvest documentation](https://rvest.tidyverse.org/)
-   [SelectorGadget](https://selectorgadget.com/) - Browser extension for finding CSS selectors
-   [W3Schools CSS Selectors](https://www.w3schools.com/cssref/css_selectors.asp)

**Ethics and legality:**

-   Always check `robots.txt` (add `/robots.txt` to any domain)
-   Respect rate limits (don't overload servers)
-   Check Terms of Service
-   Prefer APIs when available
-   Give attribution to data sources

------------------------------------------------------------------------

## Troubleshooting

**Common issues:**

-   **Error: "could not find function":** Load the packages (tidyverse, rvest)
-   **Empty or NULL results:** CSS selector might be wrong - double-check by inspecting the page
-   **Wrong number of elements:** The selector might be matching too many or too few elements
-   **Strange text with `\n` or extra spaces:** Use `str_remove_all()` and `str_squish()` to clean
-   **Can't convert to numeric:** Make sure you removed ALL non-numeric characters first (e.g., the `£` sign)
-   **Tags column looks wrong or misaligned:** Remember tags need to be pulled *quote by quote* (see Task 8.2), not all at once

**CSS selector issues:**

-   Use browser dev tools to test selectors
-   Start broad, then get more specific
-   Classes start with `.` and IDs start with `#`
-   Spaces in selectors mean "descendant" (nested elements)

**Website changes:**

-   Websites update their HTML frequently
-   If your code worked before but doesn't now, the selectors probably changed
-   Re-inspect the page and update your selectors

**If you're stuck:**

-   Check the complete solution scripts (`03-` and `04-`)
-   Use SelectorGadget to find CSS selectors
-   Ask on the discussion board
-   Attend office hours
-   Review the rvest documentation

------------------------------------------------------------------------

## Data Dictionary

### books

| Variable | Type    | Description                              |
|:---------|:--------|:-----------------------------------------|
| rank     | integer | Position on the page (1-20)              |
| title    | character | Book title (full, from the `title` attribute) |
| price    | numeric | Price in British pounds (£)              |
| rating   | integer | Star rating (1-5), decoded from the CSS class |

### quotes

| Variable            | Type      | Description                                          |
|:--------------------|:----------|:------------------------------------------------------|
| rank                | integer   | Position on the page (1-10)                          |
| quote_text          | character | The quote itself                                     |
| author              | character | Author's name                                        |
| tags                | character | Comma-separated tags for the quote                   |
| author_born_date     | character | Author's birth date - manually added for first 3 rows |
| author_born_location | character | Author's birthplace - manually added for first 3 rows |
