---
title: "AE 05 - Data Types, Factors & Dates"
author: "Your Name"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
  pdf_document:
    toc: true
    latex_engine: xelatex
    keep_tex: false
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

# Introduction

This application exercise has three short parts that all get at the same underlying question: **what kind of data is this, and how does R represent it?**

**In this application exercise, you will:**

- Predict and verify how R coerces mixed-type vectors
- Reorder the levels of a factor so they display in a meaningful order
- Tell `read_csv()` how to parse a column so it doesn't guess wrong
- Clean a numeric value that got stuck as text
- Build a real `Date` column out of separate year/month/day columns

**Estimated time:** 60-75 minutes

---

# Getting Started

## Fork and Clone Workflow

**Quick reminder:** Work in YOUR forked `application-exercises` repository.

1. Navigate to the `ae-05-data-types` folder in YOUR fork
2. Open `ae-05-data-types.Rmd`
3. Update the YAML with your name
4. Click **Knit** to verify it works

**Detailed instructions in AE 01 if needed.**

---

# Packages
```{r load-packages, warning=FALSE, message=FALSE}
library(tidyverse)
library(lubridate)
```

**What these packages do:**

- **tidyverse:** Data wrangling (dplyr), visualization (ggplot2), and reading data (readr)
- **lubridate:** Makes working with dates and times much less painful than base R

---

# Part 1: Type Coercion

R vectors can only hold **one** type of value at a time. When you combine values of different types with `c()`, R has to coerce everything into a single type behind the scenes, following this hierarchy:

$$\text{logical} \rightarrow \text{integer} \rightarrow \text{double} \rightarrow \text{character}$$

Whatever type is furthest to the right in that chain "wins."

## Exercise 1: Predict the Type

For each vector below: **first predict** what type the combined vector will be and why, **then run the code** to check, and **finally uncomment** the `typeof()` line to confirm.

**Task 1.1**
```{r ex1a}
c(1, 1L, "C")
```

**Your prediction:** _____

```{r ex1a-check}
#typeof(c(1, 1L, "C"))
```

**Task 1.2**
```{r ex1b}
c(1L / 0, "A")
```

**Hint:** Run `typeof(1L / 0)` on its own first — what does dividing an integer by zero produce?

**Your prediction:** _____

```{r ex1b-check}
#typeof(c(1L / 0, "A"))
```

**Task 1.3**
```{r ex1c}
c(1:3, 5)
```

**Your prediction:** _____

```{r ex1c-check}
#typeof(c(1:3, 5))
```

**Task 1.4**
```{r ex1d}
c(3, "3+")
```

**Your prediction:** _____

```{r ex1d-check}
#typeof(c(3, "3+"))
```

**Task 1.5**
```{r ex1e}
c(NA, TRUE)
```

**Your prediction:** _____

```{r ex1e-check}
#typeof(c(NA, TRUE))
```

**Task 1.6 (reflect):** Look back at Task 1.4. R was willing to turn `3` into `"3"` to combine it with `"3+"`, but it will never automatically go the other direction and turn `"3+"` into a number. Why not?

**Your answer:**

___________________________________________________________________________

---

# Part 2: Data Classes — Factors

## Data
```{r load-hotels-data, message=FALSE}
# From TidyTuesday: https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-02-11/readme.md
hotels <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-02-11/hotels.csv")
```

## Exercise 2: Reorder a Factor

Knit the plot below and take a look at the x-axis.

```{r ex2-plot, fig.width=10, fig.alt="Line chart comparing mean average daily rate by month for resort and city hotels, with months currently in alphabetical rather than calendar order."}
hotels %>%
  group_by(hotel, arrival_date_month) %>%
  summarise(mean_adr = mean(adr)) %>%
  ggplot(aes(
    x = arrival_date_month,
    y = mean_adr,
    group = hotel,
    color = hotel)
    ) +
  geom_line() +
  theme_minimal() +
  labs(
    x = "Arrival month",
    y = "Mean ADR (average daily rate)",
    title = "Comparison of resort and city hotel prices across months",
    subtitle = "Resort hotel prices soar in the summer while city hotel prices remain relatively constant throughout the year",
    color = "Hotel type"
    )
```

**Task 2.1:** How are the months currently ordered on the x-axis? What order would make more sense?

**Your answer:**

___________________________________________________________________________

**Task 2.2:** `arrival_date_month` is currently a character column, so R just sorts it alphabetically. Fix this by turning it into a factor with the levels in calendar order, then re-run the plot chunk above.

**Hint:** R has a built-in constant called `month.name` that's already in calendar order — look it up by running `month.name` in your Console. You'll want a function from the **forcats** package (see [forcats reference](https://forcats.tidyverse.org/reference/index.html)) to set the factor levels.

```{r ex2-fix, eval=FALSE}
# don't forget to switch eval = FALSE to TRUE after filling in the blanks
hotels <- hotels %>%
  mutate(arrival_date_month = factor(arrival_date_month, levels = ___))
```

**Task 2.3:** After re-knitting the plot with your fix, describe how the x-axis changed.

**Your answer:**

___________________________________________________________________________

**Stretch goal (optional):** The y-axis currently shows plain numbers like `80`. Change the label so values display with dollar signs, e.g. `$80`. You'll want a function from the **scales** package (see [scales reference](https://scales.r-lib.org/reference/index.html)).

```{r ex2-stretch, eval=FALSE}
# Add scale_y_continuous(labels = ___) to the plot above
```

---

# Part 3: Importing and Recoding Data

## Exercise 3: Importing Data

When you read in a file, R has to guess the type of each column — and sometimes it guesses wrong, which can silently corrupt your data.

```{r ex3-load, message=FALSE}
orders_csv <- "id,city,zip_code,price
1,Boston,02139,\"$1,200\"
2,San Francisco,94105,\"$850\"
3,New York,10001,\"$1,500\"
4,Cambridge,02138,\"$950\""

orders <- read_csv(orders_csv)
```

Run `glimpse(orders)` in your Console and look closely at the `zip_code` and `price` columns.

**Task 3.1:** What type did `read_csv()` guess for `zip_code`? Compare the printed values to the original text above — did anything change? Why is that a problem for zip codes specifically?

**Your answer:**

___________________________________________________________________________

**Task 3.2:** What type did `read_csv()` guess for `price`? Try running `mean(orders$price)` — what happens, and why?

**Your answer:**

___________________________________________________________________________

**Task 3.3:** Fix both problems. First, re-read the data telling `read_csv()` to keep `zip_code` as text using the `col_types` argument.

```{r ex3-reimport, eval=FALSE}
# don't forget to switch eval = FALSE to TRUE after filling in the blanks
orders <- read_csv(orders_csv, col_types = cols(zip_code = ___))
```

**Hint:** Look at the `col_types` examples in the R4DS Data Import reading — you want the "character" column type specifier.

Then, create a new column called `price_numeric` that strips the `$` and `,` out of `price` and converts it to a real number. Look up `parse_number()` from the **readr** package.

```{r ex3-clean-price, eval=FALSE}
# don't forget to switch eval = FALSE to TRUE after filling in the blanks
orders <- orders %>%
  mutate(price_numeric = ___(price))
```

**Task 3.4:** Verify your fix by running `mean(orders$price_numeric)`. What do you get?

**Your answer:** _____

---

## Exercise 4: Recoding Data — Dates and Times

Go back to the `hotels` data from Part 2. Right now, the booking date is split across three separate columns: `arrival_date_year`, `arrival_date_month` (a month name, like `"July"`), and `arrival_date_day_of_month`.

That's fine for grouping by month like you did in Exercise 2, but it's not useful if you wanted to know how many days apart two bookings were, or filter for a specific date range. For that, you need an actual `Date`.

**Task 4.1:** Use a function from **lubridate** to combine the three columns into one new column called `arrival_date`, containing a real `Date` value.

**Hint:** `lubridate::dmy()` can parse a string like `"1 July 2016"` directly into a date. You'll need to `paste()` the three columns together first, in day-month-year order.

```{r ex4-build-date, eval=FALSE}
# don't forget to switch eval = FALSE to TRUE after filling in the blanks
hotels <- hotels %>%
  mutate(arrival_date = dmy(paste(___, ___, ___)))
```

**Task 4.2:** Verify your fix by running `class(hotels$arrival_date)`. What does it say now?

**Your answer:** _____

**Task 4.3 (reflect):** Now that `arrival_date` is a real date instead of just a releveled factor, what could you do with the data that you couldn't easily do with the factor from Part 2?

**Your answer:**

___________________________________________________________________________
___________________________________________________________________________

---

# Reflection

**What was the trickiest part of this exercise: type coercion, factors, importing, or dates?**

___________________________________________________________________________

**How do you think the recoding/dates skills from Part 3 might come up in Lab 04?**

___________________________________________________________________________

---

# Submitting Your Work

**Before you finish:**

1. ✅ Make sure all code chunks run without errors
2. ✅ Fill in all blanks and answer all questions
3. ✅ Knit to HTML first to verify everything works
4. ✅ Review the HTML output

## Git Workflow

1. Click the **Git** pane
2. Check boxes next to all changed files
3. Click **Commit**
4. Write a commit message: "Completed AE 05 - Data Types, Factors & Dates"
5. Click **Commit**
6. Click **Push**
7. Verify your work is on GitHub in **YOUR fork**

## Generate PDF for Canvas Submission

Once everything works in HTML:

1. Click the **Knit** dropdown arrow
2. Select **Knit to PDF**
3. Wait for PDF generation
4. Find `ae-05-data-types.pdf` in the Files pane
5. Export and save to your computer
6. Submit the PDF to Canvas

**Remember:** Push your `.Rmd` to GitHub (your checkpoint/backup) AND submit the PDF to Canvas. **The PDF is the basis for grading** — make sure it's complete and readable before you submit.

---

# Variables You'll Need (Hotels Data)

| variable | class | description |
|:------------|:------------|:----------------------------------------------|
| hotel | character | Hotel (Resort Hotel or City Hotel) |
| adr | double | Average Daily Rate |
| arrival_date_year | double | Year of arrival date |
| arrival_date_month | character | Month of arrival date (as text, e.g. "July") |
| arrival_date_day_of_month | double | Day of arrival date |

For the full data dictionary, see AE 04's instructions — this is the same `hotels` dataset.

---

# Troubleshooting

**Common issues:**

- **Error: "could not find function":** Make sure you loaded the packages (`tidyverse` and `lubridate`)
- **`typeof()` line does nothing:** Make sure you actually removed the `#` in front of it
- **Factor levels don't change the plot order:** Double-check you reassigned the result back with `hotels <- hotels %>% mutate(...)` — if you don't save it back, R throws away your fix
- **`col_types` error:** Make sure you're using `col_types = cols(column_name = "c")` or `col_character()`, not just `"character"` on its own
- **`parse_number()` returns `NA`:** Check that the column still has the `$`/`,` characters — `parse_number()` strips currency symbols and thousands separators automatically, but only from character columns, not already-numeric ones
- **`dmy()` returns `NA`:** Check the order of your `paste()` arguments — it needs to be day, then month, then year, matching the function name

**If you're stuck:**

- Review the walkthrough video
- Check documentation: `?typeof`, `?factor`, `?col_types`, `?parse_number`, `?dmy`
- Ask on the discussion board
- Attend office hours

---

# Quick Reference

| Function | Package | Use |
|:---------|:--------|:----|
| `typeof()` | base R | Check the underlying type of a value or vector |
| `factor(x, levels = ...)` | base R | Turn a character vector into a factor with a specific level order |
| `col_types = cols(...)` | readr | Tell `read_csv()` exactly how to parse each column |
| `parse_number()` | readr | Pull a number out of text that has currency symbols, commas, etc. |
| `dmy()` / `mdy()` / `ymd()` | lubridate | Parse a text string into a real `Date`, based on the order of day/month/year |
