---
title: "AE 02 - Bechdel Test"
author: "Your Name"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
  pdf_document:
    toc: true
    latex_engine: xelatex
    keep_tex: false
---

## Introduction

In this mini analysis we work with the data used in the FiveThirtyEight story titled ["The Dollar-And-Cents Case Against Hollywood's Exclusion of Women"](https://fivethirtyeight.com/features/the-dollar-and-cents-case-against-hollywoods-exclusion-of-women/).

**The Bechdel Test** is a measure of the representation of women in fiction. A movie passes the test if it meets three criteria:
1. It has at least two women in it
2. Who talk to each other
3. About something other than a man

**In this application exercise, you will:**

- Practice data filtering and summarizing
- Calculate return on investment (ROI)
- Create and interpret boxplots
- Use grouping to compare categories
- Practice using inline R code
- Interpret relationships between variables

**Estimated time:** 30-45 minutes

---

## Getting Started

### Returning to `application-exercises`

You already forked and cloned this repository for AE 01 — you won't need to do that again. All of the semester's Application Exercise folders were included when you first cloned, so there's nothing new to sync or pull — the `ae-02-bechdel` folder is already sitting in your project.

1. Open your existing `application-exercises` project in RStudio
2. Navigate to the `ae-02-bechdel` folder
3. Open `ae-02-bechdel.Rmd`
4. Update the YAML with your name
5. Click **Knit** to verify it works

**Haven't forked and cloned yet?** (just joining, or missed AE 01) — see the full walkthrough in **AE 01 - UN Votes**, Steps 1-5. Everything there still applies; you just do it once.

---

## Data and Packages

We start with loading the packages we'll use.
```{r load-packages, warning=FALSE, message=FALSE}
library(fivethirtyeight)
library(tidyverse)
```

**What these packages do:**
- **fivethirtyeight:** Contains datasets from FiveThirtyEight articles
- **tidyverse:** Data wrangling and visualization

### About the Data

The dataset contains information on `r nrow(bechdel)` movies released between `r min(bechdel$year)` and `r max(bechdel$year)`. However, we'll focus our analysis on movies released between 1990 and 2013.
```{r filter-years}
bechdel90_13 <- bechdel %>% 
  filter(between(year, 1990, 2013))
```

**What this code does:**
- `filter()`: Keeps only rows meeting a condition
- `between()`: Checks if a value is between two numbers (inclusive)
- Result: Movies from 1990-2013 only

---

## Exercise 1: Count the Movies

How many movies are in our filtered dataset?

**Fill in the blank using inline code:**

There are `[your code here]` movies in our dataset.

**Hint:** Use inline R code with `nrow(bechdel90_13)` in your narrative text.

---

## Understanding the Variables

The financial variables we'll focus on are the following:

- `budget_2013`: Budget in 2013 inflation adjusted dollars
- `domgross_2013`: Domestic gross (US) in 2013 inflation adjusted dollars
- `intgross_2013`: Total International (i.e., worldwide) gross in 2013 inflation adjusted dollars

And we'll also use the `binary` and `clean_test` variables for **grouping**:

- `binary`: Whether the movie passed (PASS) or failed (FAIL) the Bechdel test
- `clean_test`: More detailed result (ok, dubious, men, notalk, nowomen)

**Explore the data:**
```{r explore-bechdel}
# View the structure
glimpse(bechdel90_13)

# See how many movies passed vs failed
bechdel90_13 %>%
  count(binary)
```

**How many movies passed the test?** [type your answer here]

**How many failed?** [type your answer here]

---

## Analysis

### Binary Bechdel Test Results

Let's take a look at how median budget and gross vary by whether the movie passed the Bechdel test, which is stored in the `binary` variable.
```{r binary-summary}
bechdel90_13 %>%
  group_by(binary) %>%
  summarise(
    med_budget = median(budget_2013),
    med_domgross = median(domgross_2013, na.rm = TRUE),
    med_intgross = median(intgross_2013, na.rm = TRUE)
  )
```

**What this code does:**
- `group_by(binary)`: Creates groups for PASS and FAIL
- `summarise()`: Calculates summary statistics for each group
- `median()`: Finds the middle value
- `na.rm = TRUE`: Removes missing values before calculating

---

## Exercise 2: Interpret Binary Results

You can answer in full sentences or bullet points — whatever helps you organize your thinking.

**Task 2.1:** Which movies have higher median budgets: those that pass or fail the Bechdel test?

**Your answer:**

Movies that [type PASS or FAIL] the test have higher median budgets.

**Task 2.2:** Which movies have higher median domestic gross: those that pass or fail?

**Your answer:**

Movies that [type PASS or FAIL] the test have higher median domestic gross.

**Task 2.3:** What might explain these patterns?

**Your answer:**

[type your answer here]

---

## Detailed Bechdel Test Results

Next, let's take a look at how median budget and gross vary by a more detailed indicator of the Bechdel test result. This information is stored in the `clean_test` variable, which takes on the following values:

- `ok` = passes test
- `dubious` = unclear whether it passes
- `men` = women only talk about men
- `notalk` = women don't talk to each other
- `nowomen` = fewer than two women

## Exercise 3: Group by Detailed Test Results

**Fill in the blank to group by `clean_test`:**
```{r detailed-summary, eval = FALSE}
bechdel90_13 %>%
  group_by(___) %>%
  summarise(
    med_budget = median(budget_2013),
    med_domgross = median(domgross_2013, na.rm = TRUE),
    med_intgross = median(intgross_2013, na.rm = TRUE)
  )
```

**Task 3.1:** Which category has the highest median budget?

**Your answer:** [type your answer here]

**Task 3.2:** Which category has the highest median international gross?

**Your answer:** [type your answer here]

---

## Return on Investment (ROI)

In order to evaluate how return on investment varies among movies that pass and fail the Bechdel test, we'll first create a new variable called `roi` as the ratio of the gross to budget.

**ROI formula:** (Total Gross) / Budget

Higher ROI = More profitable relative to what was spent
```{r create-roi}
bechdel90_13 <- bechdel90_13 %>%
  mutate(roi = (intgross_2013 + domgross_2013) / budget_2013)
```

**What this code does:**
- `mutate()`: Creates a new variable
- `(intgross_2013 + domgross_2013)`: Total gross (domestic + international)
- `/ budget_2013`: Divide by budget
- Result: ROI of 2 means the movie made twice its budget

---

## Exercise 4: Highest ROI Movies

Let's see which movies have the highest return on investment.
```{r highest-roi}
bechdel90_13 %>%
  arrange(desc(roi)) %>% 
  select(title, roi, year)
```

**What this code does:**
- `arrange(desc(roi))`: Sorts by ROI in descending order (highest first)
- `select()`: Shows only the columns we want

**Task 4.1:** What is the movie with the highest ROI?

**Your answer:** [type your answer here]

**Task 4.2:** What is its ROI value?

**Your answer:** [type your answer here]

**Task 4.3:** Does this movie surprise you? Why or why not?

**Your answer:**

[type your answer here]

---

## Visualizing ROI

Below is a visualization of the return on investment by test result. However, it's difficult to see the distributions due to a few extreme observations.

**Quick refresher — reading a boxplot:** the box shows the middle 50% of values (the interquartile range), the line inside the box is the median, the whiskers extend to the typical range of values, and individual dots beyond the whiskers are outliers.

## Exercise 5: Create Initial Boxplot

**Fill in the blank for the y-axis label:**
```{r initial-boxplot, warning=FALSE, fig.alt="Boxplot of return on investment for each detailed Bechdel test result, colored by pass/fail status, with several extreme outliers stretching the y-axis."}
ggplot(data = bechdel90_13, 
       mapping = aes(x = clean_test, y = roi, color = binary)) +
  geom_boxplot() +
  labs(
    title = "Return on investment vs. Bechdel test result",
    x = "Detailed Bechdel result",
    y = "___",#Give this axis an appropriate title
    color = "Binary Bechdel result"
  )
```

**Hint:** The y-axis shows the ROI values.

**Task 5.1:** Why is this plot difficult to interpret?

**Your answer:**

[type your answer here]

---

## Extreme Observations

What are those movies with *very* high returns on investment?
```{r extreme-roi}
bechdel90_13 %>%
  filter(roi > 400) %>%
  select(title, budget_2013, domgross_2013, year)
```

**Task 5.2:** How many movies have ROI greater than 400?

**Your answer:** [type your answer here]

**Task 5.3:** What do you notice about the budgets of these high-ROI movies?

**Your answer:**

[type your answer here]

---

## Exercise 6: Zoom In for Better View

Zooming in on the movies with `roi < 15` provides a better view of how the medians across the categories compare.

**Fill in the blanks:**
```{r zoomed-boxplot, warning=FALSE, fig.alt="Boxplot of return on investment for each detailed Bechdel test result, colored by pass/fail status, zoomed in to the 0-15 range so the medians and boxes are visible."}
ggplot(data = bechdel90_13, 
       mapping = aes(x = clean_test, y = roi, color = binary)) +
  geom_boxplot() +
  labs(
    title = "Return on investment vs. Bechdel test result",
    subtitle = "___", # Something about zooming in to a certain level
    x = "Detailed Bechdel result",
    y = "Return on investment",
    color = "Binary Bechdel result"
  ) +
  coord_cartesian(ylim = c(0, 15))
```

**Hints for subtitle:**
- Mention that you're focusing on ROI < 15
- Something like "Focusing on movies with ROI less than 15" or "Zoomed in to ROI < 15"

**What `coord_cartesian()` does:**
- Zooms in on the plot without removing data
- `ylim = c(0, 15)`: Shows only y-values from 0 to 15
- Unlike `filter()`, this doesn't remove data, just changes the view

---

## Exercise 7: Interpret the Zoomed Plot

**Task 7.1:** Looking at the zoomed plot, which category has the highest median ROI?

**Your answer:** [type your answer here]

**Task 7.2:** Do movies that pass the Bechdel test (binary = PASS) appear to have higher or lower ROI than those that fail?

**Your answer:**

[type your answer here]

**Task 7.3:** What does this suggest about the financial argument for including women in films?

**Your answer:**

[type your answer here]

---

## Exercise 8: Create Your Own Summary

**Task 8.1:** Calculate the mean ROI (not median) for movies that pass vs. fail the Bechdel test.

**Hint:** Use `group_by()` and `summarise()` like we did earlier, but with `mean()` instead of `median()`.
```{r mean-roi, eval = FALSE}
bechdel90_13 %>%
  group_by(___) %>%
  summarise(
    mean_roi = mean(___, na.rm = TRUE)
  )
```

**Mean ROI for PASS:** [type your answer here]

**Mean ROI for FAIL:** [type your answer here]

**Task 8.2:** How do these means compare to the medians we saw earlier? Why might they be different?

**Your answer:**

[type your answer here]

---

## Exercise 9: Challenge (Optional)

You've got the core skills down at this point — this last one is here for anyone who wants to stretch further. It's meant to be genuinely challenging, so don't worry if you don't finish it or need to look things up. Partial attempts are welcome.

**Create a visualization comparing ROI across years for movies that pass vs. fail the test.**

**Hints:**
- Use `geom_point()` or `geom_line()`
- Put `year` on the x-axis and `roi` on the y-axis
- Color by `binary`
- You might want to filter out extreme ROI values first
- Consider using `geom_smooth()` to show trends
- If you finish the plot, add a `fig.alt` chunk option describing what it shows — good practice for accessibility
```{r roi-over-time, warning=FALSE}
# Your code here
```

**What trend do you notice?**

[type your answer here]

---

## Knit and Submit

**Before you finish:**

1. ✅ Make sure all code chunks run without errors
2. ✅ Fill in all answer spaces
3. ✅ Knit to HTML first to verify everything works
4. ✅ Review the HTML output

**Git workflow:**

1. Click the **Git** pane
2. Check boxes next to all changed files
3. Click **Commit**
4. Write a commit message: "Completed AE 02 - Bechdel"
5. Click **Commit**
6. Click **Push**
7. Verify your work is on GitHub in **YOUR fork**

> 💬 **If only one part works:** This assignment asks for both a GitHub push and a Canvas PDF upload. If you get one working but not the other after trying the troubleshooting steps below, submit what you have by the deadline and email me — I'd rather see your progress than have you lose credit over a technical snag.

---

## Generate PDF for Canvas Submission

Once everything works in HTML:

PDF generation trips people up more than the actual R code does — if it doesn't work on the first try, that's normal, not a sign you did something wrong.

1. Click the **Knit** dropdown arrow
2. Select **Knit to PDF**
3. Wait for PDF generation
4. Find `ae-02-bechdel.pdf` in the Files pane
5. Click the checkbox next to the PDF
6. Click **More → Export...**
7. Save to your computer
8. Submit the PDF to Canvas

**Still not working after trying the troubleshooting steps below?** Knit to HTML instead, then use your browser's **Print → Save as PDF** option to create a PDF from the HTML output. It won't look identical to a true LaTeX PDF, but it's a completely acceptable submission.

**Remember:** Submit BOTH to GitHub (your .Rmd) AND Canvas (the PDF)!

---

## Reflection Questions

**Task:** After completing this exercise, reflect on what you learned. Answer in full sentences or bullet points, whichever is easier.

**What was the most interesting finding about the Bechdel test and movie finances?**

[type your answer here]

**What was the most challenging part of this exercise?**

[type your answer here]

**How did zooming in on the plot change your interpretation of the data?**

[type your answer here]

---

## References

1. Hickey, W. (2014). ["The Dollar-And-Cents Case Against Hollywood's Exclusion of Women"](https://fivethirtyeight.com/features/the-dollar-and-cents-case-against-hollywoods-exclusion-of-women/). FiveThirtyEight.

2. The Bechdel Test was popularized by Alison Bechdel's comic strip "Dykes to Watch Out For."

---

## Troubleshooting

**Common issues:**

- **Error: "could not find function":** Make sure you ran the package loading chunk
- **ROI values look wrong:** Check that you're adding domestic and international gross before dividing by budget
- **Plot doesn't show:** Make sure you ran all previous chunks that create variables
- **NA values in results:** Use `na.rm = TRUE` in your summary functions
- **Can't find the `ae-02-bechdel` folder:** Make sure you're looking inside your `application-exercises` project (not a different folder) — every AE folder for the semester was included when you first cloned it back in AE 01

**PDF issues:**

- **LaTeX not found:** Install tinytex: `tinytex::install_tinytex()`
- **PDF looks different:** This is normal - focus on content, not formatting
- **Still stuck after trying the above?** Knit to HTML and use your browser's Print → Save as PDF instead

**If you're stuck:**

- Review the walkthrough video
- Check your code against the examples
- Ask on the discussion board
- Attend office hours
